Experimental and numerical studies of the effects of resonant and random excitations of the proton beam in the LHC, with applications to the design of a pulsed hollow electron lens for active halo control

### 2018-03-01 resex.tex
Intermediate step in revision of results section on pulsing patterns.

In this Section, we present the experimental observations and compare
them with the results of numerical simulations.  In
Section~\ref{sec:pattern}, we show which pulsing patterns are most
efficient in the LHC, using the 2016 experiments as an example to show
that these patterns can be predicted.  Next, we present results for
each of the specific excitation patterns that could be tested
experimentally, namely pulsing every $10^{\mathrm{th}}$~turn
(Section~\ref{sec:simex10}), $7^{\mathrm{th}}$~turn
(Section~\ref{sec:simex7}), $8^{\mathrm{th}}$~turn
(Section~\ref{sec:simex8}), and the random excitation
(Section~\ref{sec:simexran}).  In Section~\ref{sec:damp}, we describe
how the transverse damping system influenced the effects of the
external excitation sources on losses, beam distributions, and
emittances.

\subsection{Dependence on the pulsing pattern\label{sec:pattern}}
The effect of each pulsing pattern in comparison to the other patterns can be identified by the means of the loss rate and emittance growth from the distribution tracking. As an example, the simulation results for the 2016 experiment are shown in Fig.~\ref{fig:patternsim} featuring a clear dependence of both parameters on the pattern.
\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_sigm.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit1.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_pattern_3_5um_emit2.png}
	\end{minipage}	
	\caption{\label{fig:patternsim} Relative intensity (top left), bunch length (top right) and horizontal (bottom left) and vertical (bottom right) emittance for different pulsing patterns obtained from distribution tracking based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ with standard errors. The resonant and random excitation respectively is applied in both planes with an amplitude of 96~nrad. No random noise component is added in addition.}
\end{figure*}

The largest losses are observed for $3^{\mathrm{rd}}$, $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing and a uniform random excitation. As the bunch length decreases in accordance with the losses, the losses can be mostly associated to off-momentum particles hitting the transverse aperture. Only $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing as well as random excitation exhibit considerable emittance growth. Compared to any resonant excitation, the random excitation has by far the strongest effect.

The sensitivity to $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing is also observed in absence of machine errors in which case the only source of non-linearities are sextupoles and octupoles~\cite{md_sim_hel_res_ex_fitterer} suggesting that latter are responsible for the observed sensitivity. The driven resonances are revealed by the FMA analysis shown in Fig.~\ref{fig:patternfma}. The $7Q_x$ resonance is excited in case of the $7^{\mathrm{th}}$~turn pulsing and the $10Q_x$ and $10Q_y$ resonance in case of the $10^{\mathrm{th}}$~turn pulsing. As octupoles can only drive even resonances, the source of the $7Q_x$ resonances are the sextupoles while the octupoles only generate the tune footprint. The other pulsing patterns do not exhibit any increase in losses or emittance growth without magnetic errors and their effect can be thus attributed to the magnetic field errors, implying also the seed chosen for the simulation.
\begin{figure*}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ V
		\includegraphics[width=1.0\linewidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png}
	\end{minipage}
	\caption{\label{fig:patternfma} FMA analysis without excitation (top left) for $10^{\mathrm{th}}$~turn pulsing in H+V (top left) and for $7^{\mathrm{th}}$~pulsing only in H (bottom left) and only in V (bottom right) and  based on the 2016 injection optics with no machine errors and a tune of (64.28,59.31). The excitation is 120~nrad in the corresponding plane. The absence of a strong excitation of any resonance in case of $7^{\mathrm{th}}$~turn pulsing only in V and the strong excitation in case of pulsing only in H confirms the excitation of the $7Q_x$ resonance. For $10^{\mathrm{th}}$~turn pulsing there is in contrast no significant difference between pulsing only in H, only in V or in H+V (see Appendix~\ref{app:sec:10}, Fig.~\ref{app:fig:fma:10}).}
\end{figure*}

One may have noticed that the emittance in case of the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing starts at an increased initial value and then stays almost constant during the entire simulation time, which usually points to an error in the chosen simulation setup. This behavior is however typical for any resonant excitation and can be ascribed to the adjustment of the beam distribution during the first $10^4$ turns to a new equilibrium. As the beam distribution is only dumped every $10^4$ turns, this initial fast adjustment manifests itself in an increased initial emittance value in Fig.~\ref{fig:patternsim}, and also any other simulations over $10^6$ turns presented in this paper. Besides an increased emittance, the new distribution also becomes non-Gaussian. This is illustrated by means of the residual in respect to the Gaussian distribution for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing in Fig.~\ref{fig:patternhist}. This initial fast change of the beam distribution is not an artifact of the simulation, but has been, to the knowledge of the authors, also for the first time experimentally observed during the 2016 and 2017 LHC experiments presented in this paper. The timescale of this fast adjustment is in experiments however much slower than predicted in simulations --- on the timescale of minutes compared to seconds. 
\begin{figure*}[t]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn pulsing (top left), vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (top right) and horizontal (bottom left) and vertical (bottom right) beam distribution for random excitation from distribution tracking simulations based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ and standard errors. The excitation is applied in both planes with an amplitude of 96~nrad. The residual is defined as the final distribution (here after $10^4$~turns) minus the initial distribution (here after $10^2$~turns) as \%, and the ratio as the final distribution divided by the initial distribution. The red line in the plot of the residual is the difference between the Gaussian fit of the distribution and the distribution itself.}
\end{figure*}

>>>>

In this Section, we present the experimental observations and compare
them with the results of numerical simulations.  In
Section~\ref{sec:pattern}, we show which pulsing patterns are
predicted to be the most efficient in the LHC.  Next, we present
results for each of the specific excitation patterns that could be
tested experimentally, namely pulsing every $10^{\mathrm{th}}$~turn
(Section~\ref{sec:simex10}), $7^{\mathrm{th}}$~turn
(Section~\ref{sec:simex7}), $8^{\mathrm{th}}$~turn
(Section~\ref{sec:simex8}), and the random excitation
(Section~\ref{sec:simexran}).  In Section~\ref{sec:damp}, we describe
how the transverse damping system influenced the effects of the
external excitation sources on losses, beam distributions, and
emittances.

\subsection{Dependence on the pulsing pattern}
\label{sec:pattern}

\begin{figure*}
\begin{tabular}{cc}
\includegraphics[width=0.49\textwidth]{2016injerra2b2u_pattern_3_5um_intensity.png}
  &
\includegraphics[width=0.49\textwidth]{2016injerra2b2u_pattern_3_5um_sigm.png}
  \\
\includegraphics[width=0.49\textwidth]{2016injerra2b2u_pattern_3_5um_emit1.png} &
\includegraphics[width=0.49\textwidth]{2016injerra2b2u_pattern_3_5um_emit2.png}
  \\
\end{tabular}
\caption{Relative intensity (top left), bunch length (top right) and
  horizontal (bottom left) and vertical (bottom right) emittances for
  different pulsing patterns, calculated by distribution tracking
  based on the 2016 injection optics with
  $(Q_x, Q_y) = (64.28, 59.31)$ and standard lattice errors. The
  resonant and random excitations are applied in both planes, with an
  amplitude of 96~nrad. No random noise component is added. Because of
  its much larger effects, the random excitation is shown with
  separate vertical axes.}
\label{fig:patternsim}
\end{figure*}

\begin{figure*}
  \begin{tabular}{cc}
    no excitation & $10^{\mathrm{th}}$, H+V \\
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerru_dp0_ord10_annotate.png} &
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerrut10skhv_dp0_ord10.png}
    \\
    $7^{\mathrm{th}}$, H & $7^{\mathrm{th}}$, V \\
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerrut7skh_dp0_ord7.png} &
    \includegraphics[width=0.49\textwidth]{2016injnocolc15o+19_6noerrut7skv_dp0_ord7.png} \\
  \end{tabular}
  \caption{Frequency-map analysis in betatron tune space based on the
    2016 injection optics with no machine errors and tunes (64.28,
    59.31): without excitation (top left); $10^{\mathrm{th}}$~turn
    pulsing in both horizontal and vertical planes (top left);
    $7^{\mathrm{th}}$~turn pulsing in the horizontal (bottom left) and
    vertical (bottom right) planes. The excitation amplitude is
    120~nrad in the corresponding plane. The colors (blue to red)
    represent the tune jitter of tracked particles starting at each
    given location in tune space~\cite{fmalaskar}. The absence of a
    strong excitation of any resonance in case of vertical
    $7^{\mathrm{th}}$~turn pulsing and the strong excitation in case
    of horizontal pulsing confirms the excitation of the $7 Q_x$
    resonance. For $10^{\mathrm{th}}$~turn pulsing, there is in
    contrast no significant difference between H, V, or H+V (see
    Appendix~\ref{app:sec:10}, Fig.~\ref{app:fig:fma:10}).}
  \label{fig:patternfma}
\end{figure*}

\begin{figure*}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$7^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t7skhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		$10^{\mathrm{th}}$ H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_t10skhv_3_5um_hist_y.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random H+V
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_ranhv_3_5um_hist_x.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2016injerra2b2u_3_5um_hist_y.png}
	\end{minipage}	
	\caption{\label{fig:patternhist} Horizontal beam distribution for $7^{\mathrm{th}}$~turn pulsing (top left), vertical beam distribution for $10^{\mathrm{th}}$~turn pulsing (top right) and horizontal (bottom left) and vertical (bottom right) beam distribution for random excitation from distribution tracking simulations based on the 2016 injection optics with $(Q_x,Q_y)=(64.28,59.31)$ and standard errors. The excitation is applied in both planes with an amplitude of 96~nrad. The residual is defined as the final distribution (here after $10^4$~turns) minus the initial distribution (here after $10^2$~turns) as \%, and the ratio as the final distribution divided by the initial distribution. The red line in the plot of the residual is the difference between the Gaussian fit of the distribution and the distribution itself.}
\end{figure*}

The effect of each pulsing pattern is characterized in terms of the
resulting losses and emittance growth, calculated in
distribution-tracking simulations.

As an example, the simulation results for the 2016 experiment are
shown in Fig.~\ref{fig:patternsim}.  A clear dependence of both losses
and emittances on the pattern can be seen.  The largest losses are
predicted for $3^{\mathrm{rd}}$, $7^{\mathrm{th}}$ and
$10^{\mathrm{th}}$~turn pulsing, and for a uniform random
excitation. Because the bunch length decreases with the number of lost
particles, losses are attributed to off-momentum particles hitting the
transverse aperture. Significant emittance growth is visible only for
$7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing and for the
random excitation. Compared to any resonant excitation, the random
excitation shows by far the strongest effect.

The effects of $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing
are also observed in the absence of machine lattice errors. In this
case, the only sources of nonlinearity are sextupoles and
octupoles~\cite{md_sim_hel_res_ex_fitterer}, suggesting that these
nonlinearities are responsible for the pronounced beam
sensitivity. The driven resonances are revealed by the frequency-map
analysis, shown in Fig.~\ref{fig:patternfma}. The $7 Q_x$ resonance is
excited by the $7^{\mathrm{th}}$~turn pulsing, and the $10 Q_x$ and
$10 Q_y$ resonances by the $10^{\mathrm{th}}$~turn pulsing. As
octupoles can only drive even resonances, the sources of the $7 Q_x$
resonances are the sextupoles, while the octupoles generate the tune
footprint. The other pulsing patterns do not exhibit any increase in
losses or emittance growth without magnetic errors. Their effect can
be thus attributed to an interaction of the excitation with the
lattice errors, implying also a sensitivity to the chosen random seed
in simulations and to the specific lattice in experiments.

One may have noticed that the emittance in case of the
$7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing starts at an
increased initial value and then stays almost constant during the
entire simulation time, which usually points to an error in the chosen
simulation setup. This behavior is however typical for any resonant
excitation and can be ascribed to the adjustment of the beam
distribution during the first $10^4$ turns to a new equilibrium. As
the beam distribution is only dumped every $10^4$ turns, this initial
fast adjustment manifests itself in an increased initial emittance
value in Fig.~\ref{fig:patternsim}, and also any other simulations
over $10^6$ turns presented in this paper. Besides an increased
emittance, the new distribution also becomes non-Gaussian. This is
illustrated by means of the residual in respect to the Gaussian
distribution for $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing
in Fig.~\ref{fig:patternhist}. This initial fast change of the beam
distribution is not an artifact of the simulation, but has been, to
the knowledge of the authors, also for the first time experimentally
observed during the 2016 and 2017 LHC experiments presented in this
paper. The timescale of this fast adjustment is in experiments however
much slower than predicted in simulations --- on the timescale of
minutes compared to seconds.
